Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-29 ind_files/04_field_eqns.tex
Rephrase after proofreading

In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$
\begin{equation}
\label{functional_derivative_defn}
    \int\frac{\delta F}{\delta \phi(x)}\rho(x)dx = \lim_{h \to 0} \frac{F[\phi + h\rho]-F[\phi]}{h}
\end{equation}
This definition is formulated in analogy to directional derivatives in multi-variable calculus: Noting that a function can be thought of as an infinite-dimensional vector, informally `taking the limit' $n \to \infty$ in \eqref{directional_derivative_defn} yields \eqref{functional_derivative_defn}.

@@

for every `nice' subset $A \subset \mathcal{T}$. The Dirac mass is often written $\delta_{x_i} = \delta(y-x_i)$ as a `function' of a dummy variable $y$ which will be integrated over, since in this view the Dirac mass is a `function' that can only occur inside an integral. Though I will indeed (kind of) use the Dirac mass as a function, I will stick to the notation $\delta_{x_i}$ because it emphasizes that $\delta_{x_i}$ is supposed to represent an individual with a trait value of $x_i$ (the dummy variable $y$ can be confusing in this regard). Note that by choosing $f(x) \equiv 1$, we get an `indicator' that is $1$ if the individual is within the set $A$ and $0$ otherwise. If the population at any time $t$ consists of $N(t)$ individuals with trait values $\{x_1,x_2,\ldots,x_{N(t)}\}$, then it can be completely characterized (Figure \ref{fig_infD_pop_description}) by the `distribution'

@@

\section{The infinite population limit}
Taking $K \to \infty$ in equation \eqref{functional_langevin} yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where I have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$ in which each $\phi^{(t)}$ is a finite measure. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE models' discussed in studies of Adaptive Diversification~\citep{doebeli_adaptive_2011}. Equation \eqref{deterministic_traj} is also the basic equation of `oligomorphic dynamics'~\citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology~\citep{nordbotten_dynamics_2020}. A prominent recent example is~\posscite{wickman_theoretical_2022} `trait space equations' in their framework for eco-evolutionary community dynamics.

@@

Equation \eqref{cts_replicator_mutator} also recovers~\posscite{kimura_stochastic_1965} continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming\footnote{This actually need not be an assumption, and can be derived as a theorem under biologically reasonable conditions on the birth/death functionals upon taking the rigorous approach to both the system size expansion (rescaling) and the infinite population limit. See, for example,~\citet{etheridge_introduction_2000} or proposition 4.1 in~\citet{champagnat_individual_2008}} that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have $m(y-z)\psi(z,t) \in  \mathcal{L}^{1}(\mathbb{R}\times\mathbb{R})$ and we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

@@

Note that by definition, $B(y(t))$ is the statistical variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Equation \eqref{AD_canonical_derived} is the canonical form of a broad class of systems called `gradient equations' or `gradient dynamics' in quantitative genetics~\citep{lande_quantitative_1982, abrams_relationship_1993, lehtonen_price_2018, lion_theoretical_2018}, and captures the approximate evolutionary dynamics of quantitative traits under certain mutation limits.  It is also deeply related~\citep{lehtonen_price_2018, lion_theoretical_2018} to the canonical equation of adaptive dynamics first formulated in~\citet{dieckmann_dynamical_1996}. The major difference is that in the `proper' canonical equation of adaptive dynamics (as formulated in~\citet{dieckmann_dynamical_1996}), the function $B(y)$ explicitly relies on mutations as a continual source of variation, whereas in gradient dynamics and our equation \eqref{AD_canonical_derived}, $B(y)$ captures the standing genetic variation in the population but does not specify the source of this variation. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are vanishingly rare and mutants are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved rigorously using much more sophisticated mathematical tools grounded in martingale theory~\citep{champagnat_unifying_2006}. A heuristic derivation of the canonical equation of adaptive dynamics is provided in the classic article by~\citet{dieckmann_dynamical_1996}.

>>>>

In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$
\begin{equation}
\label{functional_derivative_defn}
    \int\frac{\delta F}{\delta \phi(x)}\rho(x)dx = \lim_{h \to 0} \frac{F[\phi + h\rho]-F[\phi]}{h}
\end{equation}
This definition is formulated in analogy to directional derivatives in multi-variable calculus: Noting that a function can be thought of as an infinite-dimensional vector, informally `taking the limit' $m \to \infty$ in \eqref{directional_derivative_defn} yields \eqref{functional_derivative_defn}.

@@

for every subset $A \subset \mathcal{T}$. The Dirac mass is often written $\delta_{x_i} = \delta(y-x_i)$ as a `function' of a dummy variable $y$ which will be integrated over, since in this view the Dirac mass is a `function' that can only occur inside an integral. Though I will indeed (kind of) use the Dirac mass as a function, I will stick to the notation $\delta_{x_i}$ because it emphasizes that $\delta_{x_i}$ is supposed to represent an individual with a trait value of $x_i$ (the dummy variable $y$ can be confusing in this regard). Note that by choosing $f(x) \equiv 1$, we get an `indicator' that is $1$ if the individual is within the set $A$ and $0$ otherwise. If the population at any time $t$ consists of $N(t)$ individuals with trait values $\{x_1,x_2,\ldots,x_{N(t)}\}$, then it can be completely characterized (Figure \ref{fig_infD_pop_description}) by the `distribution'

@@

\section{The infinite population limit}
Taking $K \to \infty$ in equation \eqref{functional_langevin} yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where I have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$ in which each $\phi^{(t)}$ is a finite measure. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE models' discussed in studies of Adaptive Diversification~\citep{doebeli_adaptive_2011}. Equation \eqref{deterministic_traj} is also the basic equation of `oligomorphic dynamics'~\citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}(t)\psi_k(x,t)$, where $n_{k} \geq 0$ is the abundance of the $k$\textsuperscript{th} morph, $\psi_k(x,t)$ is the phenotypic distribution of the $k$\textsuperscript{th} morph (often assumed a normal distribution with narrow variance) and $S$ is the total number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology~\citep{nordbotten_dynamics_2020}. A prominent recent example is~\posscite{wickman_theoretical_2022} `trait space equations' in their framework for eco-evolutionary community dynamics.

@@

Equation \eqref{cts_replicator_mutator} also recovers~\posscite{kimura_stochastic_1965} continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming\footnote{This actually need not be an assumption, and can be derived as a theorem under biologically reasonable conditions on the birth/death functionals upon taking the rigorous approach to both the system size expansion (rescaling) and the infinite population limit. See, for example,~\citet{etheridge_introduction_2000} or proposition 4.1 in~\citet{champagnat_individual_2008}} that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have
\begin{equation*}
\int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}}|m(y-z)\psi(z,t)|dydz =  \int\limits_{\mathbb{R}}\left(\int\limits_{\mathbb{R}}|m(y-z)|dy\right)|\psi(z,t)|dz
= \|m\|_{L^{1}(\mathbb{R})}\|\psi\|_{L^{1}(\mathbb{R})} < \infty
\end{equation*}
We can therefore conclude that $m(y-z)\psi(z,t) \in  L^{1}(\mathbb{R}\times\mathbb{R})$ for any given $t>0$, meaning we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

@@

Note that by the definition of statistical variance, the function $B(y(t))$ is numerically equal to $\sigma^2_{x}(t)$, the statistical variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Equation \eqref{AD_canonical_derived} is the canonical form of a broad class of systems called `gradient equations' or `gradient dynamics' in quantitative genetics~\citep{lande_quantitative_1982, abrams_relationship_1993, lehtonen_price_2018, lion_theoretical_2018}, and captures the approximate evolutionary dynamics of quantitative traits under certain mutation limits.  It is also deeply related~\citep{lehtonen_price_2018, lion_theoretical_2018} to the canonical equation of adaptive dynamics first formulated in~\citet{dieckmann_dynamical_1996}. The major difference is that in the `proper' canonical equation of adaptive dynamics (as formulated in~\citet{dieckmann_dynamical_1996}), the function $B(y)$ explicitly relies on mutations as a continual source of variation, whereas in gradient dynamics and our equation \eqref{AD_canonical_derived}, $B(y)=\sigma^2_{x}(t)$ captures the standing genetic variation in the population but does not specify the source of this variation. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are vanishingly rare and mutants are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved rigorously using much more sophisticated mathematical tools grounded in martingale theory~\citep{champagnat_unifying_2006}. A heuristic derivation of the canonical equation of adaptive dynamics is provided in the classic article by~\citet{dieckmann_dynamical_1996}.
